---
title: "homework-03"
format: 
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
author: "Sumaiya Alvi"
date: "2026-02-26"
---

[Link to Github repository](https://github.com/sumaiyaalvi/ENVS-193DS_homework-03.git)

[Deployed html url](https://sumaiyaalvi.github.io/ENVS-193DS_homework-03/code/homework-03.html)

# Set up tasks

```{r}
#| label: load packages and data
library(tidyverse) #general use
library(here) #file organization
library(janitor) #cleaning data frames
library(readxl) #reading excel files


salinity <- read_csv(
  here( #allows you to write in file paths
    "data", #folder
    "salinity-pickleweed.csv" #salinity data
  ))

screentime <- read_csv(
  here( 
    "data", #data folder
    "screentime.csv" #personal data 
  ))
```
# Problem 1: Slough soil salinity

## a. An appropriate test

The appropriate tests to determine the strength of the relationship between California pickleweed biomass and soil salinity are Pearson’s r correlation and simple linear regression, since both biomass (measured in grams) and soil salinity (measured as electrical conductivity in mS/cm) are continuous variables. Pearson’s r measures the strength and direction of the linear association between salinity and biomass, and quantifies how closely changes in salinity are related to changes in plant biomass. In contrast, simple linear regression treats biomass as the response variable and salinity as the predictor, estimating how much biomass changes for each one-unit increase in salinity, and using $R^2$ to quantify how much of the variation in biomass is explained by salinity.

## b. Create a visualization

```{r}
#| label: salinity and biomass scatter plot

#base layer: ggplot
ggplot(data = salinity,
       aes(x = salinity_mS_cm,
           y = pickleweed)
       ) +
  #first layer: adding scatter plot points to represent individual pickleweed plants
  geom_point(color = "#3b7ea1") +
  #second layer: adding a linear regression line to show the trend
  geom_smooth(method = "lm", color = "#1b2a41", se = TRUE) +
  #labelling the x and y axes
  labs(
    title = "As soil salinity increases, pickleweed biomass also increases",
    x = "Soil salinity (mS/cm)",
    y = "California pickleweed biomass (g)"
  ) +
  #changing the theme from the ggplot default
  theme_classic()
```

## c. Check your assumptions and run test

### Checking assumptions for simple linear regression

**Assumption 1: Linearity**

Linearity was assessed previously using the scatterplot with a fitted regression line in 1b. That plot shows a roughly straight-line trend, so the linearity assumption has been met. 

**Assumption 2: Normality of residuals**

```{r}
#fitting linear regression model: biomass predicted by salinity
sal_lm <- lm(pickleweed ~ salinity_mS_cm, data = salinity)

#creating Q-Q plot of residuals
ggplot(data = data.frame(resid = resid(sal_lm)),
       aes(sample = resid)) +
  geom_qq(color = "#3b7ea1") +
  geom_qq_line() +
  labs(
    title = "Q-Q plot of regression residuals",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  theme_light()
```

```{r}
#Shapiro-Wilk test for normality of residuals
shapiro.test(resid(sal_lm))
```
**Assumption 3: Homoscedasticity (constant variance)**

```{r}
#plotting residuals vs fitted values to assess constant variance
ggplot(data = data.frame(fitted = fitted(sal_lm),
                         resid = resid(sal_lm)),
       aes(x = fitted, y = resid)) +
  geom_point(color = "#3b7ea1") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs fitted values",
    x = "Fitted biomass (g)",
    y = "Residuals"
  ) +
  theme_light()
```

For the simple linear regression model showing the relationship between soil salinity (mS/cm) and pickleweed biomass (g), I checked the assumptions of linearity (previously assessed in the scatterplot in 1b), normality of residuals, and homoscedasticity. I evaluated normality using a Q–Q plot and a Shapiro–Wilk test of the residuals (p = 0.31), and I assessed constant variance using a residuals vs. fitted values plot. The linearity assumption was met, the Q–Q plot shows points falling approximately along the reference line, the Shapiro–Wilk test is not significant, and the residual plot shows no clear pattern or funnel shape, so the assumptions for simple linear regression appear reasonably met.

### Running the test (simple linear regression)

```{r}
#displaying regression results
summary(sal_lm)
```

## d. Results

To evaluate the strength of the relationship between California pickleweed biomass (g) and soil salinity (mS/cm), I used a simple linear regression because both variables are continuous and I wanted to quantify how biomass changes with increasing salinity. The results show a significant positive relationship, with biomass increasing by about 2.08 g for every 1 mS/cm increase in salinity (linear regression: slope = 2.08, $R^2$ = 0.29, p = 0.0086). This indicates that higher soil salinity is associated with greater pickleweed biomass, and salinity explains about 29% of the variation in biomass in this study.





